Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-03 Manuscript/manuscriptPS.tex

We used the headgroup C--H bond order parameters, $S_{\rm CH}$, and the open collaboration approach to evaluate the quality
of the headgroup structure and the ion binding affinity 
in available MD models of PS lipid.
The main advantage of this approach is the direct connection
between the order parameters measured accurately in experiments and calculated readily from simulations. 
This reduces the ambiguity in the interpretation of experiments.

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
and then proceeded to compare MD simulation results from several force fields with the experimental data.
This revealed that none of the force fields
reproduce the PS headgroup order parameters within the experimental accuracy.
However, the best models captured essential differences with PC and
suggested characteristic conformations of PS headgroups.
Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers with two exceptions: 1) the MacRog simulation with potassium counterions appears to produce the most realistic monovalent ion binding
affinity to PS-containing lipid bilayers, and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16} underestimates the calcium binding affinity.
The experimentally measured response of the PS headgroup order parameters to the bound calcium, as well as to the dilution of bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields. This underlines the need for more accurate
MD force fields to study the biological function of PS lipids.
This is in contrast to the previous results with PC lipids,
where the experimentally measured headgroup order parameter responses to the bound charge
were in qualitative agreement even though the headgroup structures themselves were
incorrect and the cation binding affinities were overestimated \cite{catte16}.

We expect our results to pave the way for the development of better MD force fields for PS lipids.
The quality of conformational ensembles produced by the various models can be evaluated
based on the headgroup order parameters, which we hope to guide the development of force fields towards
models that correctly reproduce the PS headgroup structures. Also, the ensembles observed in the simulations
with the most realistic headgroup conformations (CHARMM36 and Slipids) hint the direction for the force
field improvement. The cation binding can be improved based on the experimental headgroup order parameter
data from POPC:POPS (5:1) mixtures under different cation concentrations. Recent studies showed that the
cation binding to bilayers with POPC and POPS lipids can be improved by implicit inclusion of electronic
polarizability using the electronic continuum correction (ECC) \cite{melcr18,ECCpops}, suggesting that
the electronic polarizability effects are important for lipid-ion interactions but polarizable force fields
may not be necessary to correctly capture the ion binding to lipid bilayers.

>>>>

We used the headgroup C--H bond order parameters, $S_{\rm CH}$, and the open collaboration approach to evaluate the quality
of the headgroup structure and the ion binding affinity 
in available MD models of PS lipids.
The main advantage of using the $S_\mathrm{CH}$ is the direct connection they provide
between experiments and simulations: They can be measured accurately in experiments and calculated readily from simulations. 
This reduces the ambiguity in the interpretation of experiments.

First, we complemented the available experimental information \cite{browning80,roux90} by measuring the signs of the PS headgroup order parameters,
and then proceeded to compare MD simulation results from several force fields with the experimental data.
This revealed that none of the force fields
reproduce the PS headgroup order parameters within the experimental accuracy.
However, the best models captured essential differences between PS and PC, and
suggested characteristic conformations of PS headgroups.
Comparison to the experimentally observed order parameters in POPC:POPS (5:1) bilayers  at varying ion
concentrations \cite{roux90} then showed that the tested MD force fields
overestimate the cation binding affinity to these bilayers. There were two exceptions: 1) the MacRog force field with potassium counterions, which appeared to display a more realistic monovalent ion binding
affinity to PS-containing lipid bilayers than the other models; and 2) the CHARMM36 force field with the recently introduced
NBfix correction for calcium \cite{kim16}, which underestimated the calcium binding affinity.
Importantly, the experimentally measured responses of the PS headgroup $S_\mathrm{CH}$ to bound calcium, as well as to dilution of the bilayer with zwitterionic PC lipids, were not
qualitatively reproduced in any of the tested force fields.
This is in contrast to previous results with PC lipids,
where MD force fields were seen to be at least in qualitative agreement with the experimentally measured headgroup $S_\mathrm{CH}$ responses to bound charge---even
though the headgroup structures themselves were
incorrect and the cation binding affinities were overestimated \cite{catte16}.
This underlines the dire need for more realistic MD force fields to study the biological function of PS lipids.

We expect our results to pave the way for the development of better PS force fields.
As the quality of the conformational ensembles can be evaluated
against the headgroup C--H order parameters, we hope that the $S_\mathrm{CH}$ will guide the development of
models that correctly reproduce the PS headgroup structures. For example, the ensembles we observed in the simulations
with the most realistic headgroup conformations (CHARMM36 and Slipids) do hint a direction for force
field improvement. The cation binding could be improved based on the experimental headgroup $S_\mathrm{CH}$
data from POPC:POPS (5:1) mixtures under different cation concentrations. Recent studies have shown that 
cation binding to bilayers with POPC and POPS lipids can be improved by implicit inclusion of electronic
polarizability using the electronic continuum correction (ECC) \cite{melcr18,ECCpops}, suggesting that
the electronic polarizability effects are important for lipid--ion interactions but polarizable force fields
may not be necessary to correctly capture ion binding to lipid bilayers.
